Structuring Theories with Implicit Morphisms

### 2018-09-26 intro.tex
rewritten Sect. 1+2

The contribution of this paper is threefold.

Firstly, we describe the syntax and semantics of include declarations in \mmt.
Our main interest here is that inclusion are the most important form of implicit morphisms.
However, this is also the first item that the semantics of includes in MMT is specified rigorously.

Secondly, we generalize inclusions to implicit morphisms and show how we can use implicit morphisms to structure large diagrams of theories.
Most notably, we show that many of the goals of realms \cite{realms} (which won the best paper award at MKM 2014) are already realized by \mmt in the presence of implicit morphisms.

Thirdly, we give an extension of \mmt with complex theory and morphism expressions.
Our focus here is on defining which of these complex morphisms can be designated as implicit.
However, the language is novel and a valuable contribution in itself.

>>>>

\paragraph{Motivation}
Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps}, and knowledge representation \cite{RK:mmt:10}.

%The latter of these tools is our \mmt system.
%We use \mmt as both a theoretical framework and a concrete implementation of our results.
%But our ideas are independent of \mmt and can be transferred very easily to any other language using theory morphisms.

These systems usually use a logic $L$ for the fine-granular formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
This diagram is generated by all theories/morphisms defined, induced, or referenced in a user's development.

\begin{wrapfigure}{r}{5.5cm}
\vspace{-3em}
\begin{tikzpicture}
\node (M) at (0,0) {$\cn{Monoid}$};
\node (G) at (-2,-2) {$\cn{Group}$};
\node (D) at (2,-2) {$\cn{DivGroup}$};
\draw[mono](M) -- (G);
\draw[arrow](G) to[out=30,in=150] node[above] {$\cn{G2DG}$} (D);
\draw[arrow](D) to[out=-150,in=-30] node[above] {$\cn{DG2G}$} (G);
\end{tikzpicture}
\vspace{-3.5em}
\end{wrapfigure}

For example, a user might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
This would result in the diagram on the right.
Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$.
Crucially, $m(-)$ preserves typing and provability.

The key idea behind implicit morphisms is very simple:
We maintain an additional diagram $I$, which is commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipc{i}{S}{T}$.
Commutativity makes the following language extension well-defined: if $\ipc{i}{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions; and if $c$ is used in a $T$-expression, the semantics of $c$ is $i(c)$ where $i$ is the uniquely determined implicit morphism $i:S\to T$.

Despite their simplicity, the practical implications of implicit morphism are huge.
For example, in the diagram above, we may choose to label $\cn{G2DG}$ implicit.
Immediately, every abbreviation or theorem that we have formulated in the theory $\cn{Group}$ becomes available for use in $\cn{DivGroup}$ without any syntactic overhead.
We can even label $\cn{DG2G}$ implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that $\cn{Group}$ and $\cn{DivGroup}$ are just different formalizations of the same concept.
While these morphisms must be labeled manually, any inclusion morphism like the one from $\cn{Monoid}$ to $\cn{Group}$ is implicit automatically.

\paragraph{Contribution}
We present a formal system for developing structured theories with implicit morphisms.
Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
\mmt is foundation-independent in the sense that it allows embedding a large variety of declarative languages (logics, type-theories, etc.).
Therefore, all our results are foundation-independent --- users of a particular language $L$ can use \mmt directly as a structuring formalism for $L$ or can easily transfer our results to a dedicated implementation of $L$.

More precisely, out contribution is twofold.

Firstly, we give a novel presentation of the syntax and semantics \mmt.
Contrary to the original presentation in \cite{RK:mmt:10}, which hard-coded a single structuring mechanism --- there called \textit{structures} --- our definitions allows for adding arbitrary independent structuring mechanisms.
This required a from-scratch redesign of the original presentation in \cite{RK:mmt:10} (which is why this paper has no preliminaries section).

Based on this extensible definition, we can introduce several additional structuring mechanisms.
Many of these have already been implemented in the \mmt tool but were not formally defined  a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
With our reformulation, this is now not only possible but very easy and elegant.
As an example, we give includes and unions of theories, which induce the most important special cases of implicit morphisms.

Secondly, we describe implicit morphisms as a novel feature of \mmt and give a number of examples.
In particular, we recover the concept of realms \cite{realms} as a special case of implicit morphisms.

In fact, implicit morphisms work so well that we have refactored the \mmt tool in such a way that implicit morphisms are now more primitive than inclusion morphisms.
The semantics of inclusion morphisms is obtained by saying that inclusions are implicit morphisms that map all identifiers to themselves.
Even the fundamental property that a theory may reference its own identifiers is then just a consequence of the fact that all identity morphisms are implicit.
Therefore, surprisingly, adding the new feature of implicit morphisms to the \mmt kernel has made its design much simpler.

### 2018-09-26 mmt.tex
rewritten Sect. 1+2

The \mmt language \cite{RK:mmt:10} provides an extremely general setting in which we can define theories.
The precise definitions of the well-formed theories $T$ and theory morphisms $M$ on the one hand and their semantics $\flt{T}$ and $\flt{M}$ on the other hand are \textbf{mutually recursive}.
Therefore, we first give an overview and then recap the definitions below.

\textbf{Flat theories} are lists of symbol declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
\textbf{Flat morphisms} from a theory $S$ to a theory $T$ are lists of symbol assignments $c:=e$ where $c$ is declared in $S$ and $e$ is an expression over $T$.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
Every morphism $M$ induces a \textbf{homomorphic extension} $M(-):\Exp{S}\to\Exp{T}$, which replaces every $c\in\dom{S}$ in an $S$-expressions with the $T$-expressions $o$ such that $(c:=o) \in \flt{M}$.
Both theories and morphisms are subject to typing conditions that are not essential for our purposes here.

Flat theories and morphisms are rarely constructed in practice --- instead, users often construct larger theories from smaller ones.
We distinguish two kinds of constructions:
\begin{compactitem}
 \item \textbf{Structured theories} contain additional declarations such imports and instantiations or high-level declarations.
 \item \textbf{Theory expressions} are operators that take (normal or theory) expressions as input and return theory expressions.
\end{compactitem}
This distinction is orthogonal to the semantic variety, and both kinds are equally expressive, e.g., we can construct the union of two theories using either construction.
For example, we can construct the theory of commutative groups as the structured theory containing the declarations $\icl{Group},\icl{CommutativeMagma}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
All of the above applies analogously to structured \emph{morphisms} and \emph{morphism} expressions.

Despite the syntactic difference, the semantics of both kinds of constructions can be defined by uniformly by mapping structured theories/theory expressions $T$ to flat theories $\flt{T}$.
We call this mapping \textbf{flattening}.

An MMT \textbf{diagram} consists of a set of structured theory/morphism declarations and induces sets of theory and morphism expressions.
The names of the former are the \textbf{atomic theories/morphisms}, and they provide the base case of the inductive construction of the latter.
Vice versa, the latter may already occur in the bodies of the structured declarations.
Thus, the two concepts must be defined by mutual recursion.

More formally, every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of atomic theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
Then flattening assigns
\begin{compactitem}
 \item to each $T\in\Thy$ the flat theory $\flt{T}$ (seen as a set of declarations),
 \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$ (seen as a set of assignments).
\end{compactitem}

Within this framework, we can define new kinds of structuring declarations and new kinds of expression-forming operators along with their flattening.
In both cases, the semantics should be \textbf{compositional} in the following sense:
The flattening of a structured theory arises as the union of the flattenings of each declaration in it.
More formally, for a list $D$ of declarations and a declaration $d$, we have $\flt{D,d}=\flt{D}\cup\fltd{\flt{D}}{d}$ where $\fltd{T}{d}$ defines the semantics of declaration $d$ in the context of theory $T$.
Similarly, the flattening of a theory expression such as $f(S,T)$ depends only on the flattenings of its arguments
More formally, e.g., for theory expressions $S$ and $T$ and a binary theory-forming operator $f$, we have $(\flt{f(S,T)}=\flt{f}(\flt{S},\flt{T})$ where $\flt{f}$ is a function defining the semantics of $f$.

@@

We use the non-modular variant of \mmt, which does not use imports and instantiations between theories.
The full version can be found in \cite{RK:mmt:10}.

\begin{definition}[Theory]\label{def:theory}
A \textbf{theory} $T$ is of the form
\begin{grammar}
TDec     & t=\{Dec,\ldots,Dec\}  & \text{atomic theory declaration} \\
Dec      & n\opt{:o}\opt{=o}     & \text{symbol declaration}\\
c        & t?n                   & \text{qualified symbol identifiers} \\
o        & c \alt \ldots         & \text{OpenMath objects built from symbols $c$} \\
T        & t                     & \text{theories} 
\end{grammar}

In an atomic theory, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed OpenMath objects using only previously introduced symbols.
We omit the remaining productions for objects here.
\end{definition}

These theories are similar to OpenMath content dictionaries \cite{openmath}.
In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
But the interface always knows the qualified identifier and shows it, e.g., on hover.}
Thus, symbols of the same name declared in different atomic theories are always distinguished.

Symbol declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
Similarly, objects subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

Individual formal languages arise as fragments of \mmt: They single out the well-formed objects by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The details can be found in \cite{rabe:howto:14}.

\begin{definition}[Morphism]\label{def:morphism}
For two theories $S$ and $T$, a \textbf{morphism} $M$ is of the form
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{atomic morphism declaration}\\
Ass    & c:=o                            & \text{assignment to symbol}\\
M      & m \alt \id{T} \alt M;M          & \text{morphisms} 
\end{grammar}
such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
\end{definition}

Crucially, morphisms \textbf{preserve judgments}, in particular they map theorems to theorems.
Formally, an atomic morphism $m:S\to T$ is \emph{well-formed} if
\begin{compactitem}
\item whenever $c$ has type $t$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o:m(t)$,
\item whenever $c$ has definiens $d$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o\equiv m(d)$.
\end{compactitem}
Then we can show that for every $M\in\Mo{S}{T}$ if $\der_S o:o'$ (or $\der_S o\equiv o'$), then $\der_T M(o):M(o')$ (or $\der_T M(o)\equiv M(o')$).
The details can be found in \cite{rabe:howto:14}.
%In particular, this means that $M$ maps $S$-theorems to $T$-theorems.

\begin{definition}[Diagram]
A \textbf{diagram} is a list of atomic theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
A diagram is well-formed if each atomic theory/morphism declaration is well-formed relative to the diagram preceding it.
\end{definition}

@@

\paragraph{Induced Declarations and Assignments}
For every atomic theory $t=\{\Sigma\}$, we have $t\in\Thy$.
We define $\flt{t}$ by induction on the declarations in $\Sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\Sigma,\;n[:t][=d])}=\flt{\Sigma}\cup\{t?n[:t][=d]\}\]
where $\cdot$ is the empty sequence.
Correspondingly, for every atomic morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;c:=o)}=\flt{\sigma}\cup\{c:=o\}\]

For the remaining morphisms, the semantics is not as trivial but still straightforward.
For the identity, we have $\id{T}\in\Mo{T}{T}$ if $S\in\Thy$, and $\flt{\id{T}}=\{c:=c|c\in\dom{T}\}$.
For the composition, we have $M;N\in\Mo{R}{T}$ if $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, and $\flt{(M;N)}=\{c:=N(M(c))|c\in\dom{R}\}$.

\paragraph{Equality and Inclusion}
We say that $S$ is \textbf{included} into $T$, written $S\harr T$, if $\flt{S}\sq\flt{T}$.
%In that case, there is an inclusion morphism $S\to T$ (which we also denote $S\harr T$) defined by $\flt{m}=\flt{\id{S}}$.
%Clearly, $m(o)=o$ for all $S$-objects $o$.
Thus, every declaration of $S$ is also available in $T$ and $\Exp{S}\sq\Exp{T}$.
This induces an inclusion morphism, which maps every $c\in\dom{S}$ to itself.
We do not introduce a name for that morphism and simply define $\id{S}\in\Mo{S}{T}$ whenever $S\harr T$ (which includes the special case $S=T$).
Moreover, if $S\harr T$ and $M\in\Mo{T}{U}$, we define the restriction $M|_S\in\Mo{S}{U}$ of $M$ to $S$ by $\id{S};M$.

@@

Correspondingly, we say that two morphisms $M,N\in\Mo{S}{T}$ are \textbf{equal}, written $M\equiv N$, if $\flt{M}=\flt{N}$.
In that case, $\der_T M(o)\equiv N(o)$ for all well-formed $S$-objects $o\in\Exp{S}$.

Clearly, $\equiv$ is an equivalence relation and $\harr$ is an order relation (if anti-symmetry is taken with respect to $\equiv$).
Identity and composition are congruent with respect to $\equiv$.
Moreover, composition is associative and identity neutral with respect to $\equiv$.
Thus, each diagram yields the structure of a category.

\paragraph{Preview}
The above recap of (the non-modular fragment of) \mmt deviates from \cite{RK:mmt:10}, very carefully wording the definitions in a way that makes stating the results of this paper most elegant.
Thus, we intentionally introduced some peculiarities:
\begin{compactitem}
\item We have $\Thy=\thy$ because there are no non-atomic theories.
We will introduce some in Sect.~\ref{sec:complex}.
\item The definitions of $\flt{t}$ and $\flt{m}$ are overly complicated.
That will pay off when we introduce additional declarations in Sect.~\ref{sec:include}: We can define their semantics simply by adding a case to the inductions in $\flt{\Sigma}$ and $\flt{\sigma}$.
\item $s\harr t$ never holds because all $c\in\dom{s}$ are of the form $s?n$ and all $c\in\dom{t}$ are of the form $t?n$.
In fact, $\flt{S}$ and $\flt{T}$ are disjoint whenever $S\neq T$.
That will change when we introduce include declarations in Sect.~\ref{sec:include}.
\end{compactitem}

>>>>

\subsection{Overview}

\paragraph{Modularity in MMT}
As much as possible, theories and morphisms are treated uniformly, and we use the word \textbf{module} to refer to them collectively.
\mmt allows constructing large modules from small ones in two ways:
\begin{compactitem}
  \item \textbf{structured modules} contain declarations that import other, previously defined ones,
  \item \textbf{module expressions} are anonymous expressions (akin to formulas, terms, etc.) that denote modules.
\end{compactitem}

These two kinds of constructions have very different advantages but the same expressivity.
For example, we can construct the theory of commutative groups as the structured theory $\cn{CommutativeGroup}=\{\icl{Group},\icl{CommutativeMagma}\}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
The former is more practical when building named toplevel theories in software systems because each include is a separate declaration that can be parsed, checked, and flattened individually.
While negligible in this example, this becomes relevant quickly in large theories with complex structure.
The latter is more practical when building anonymous theory that are only needed temporarily.
In software systems, they also allow reusing operations like equality and $\lambda$-abstraction that already be implemented at the level of the base logic.
Building large morphism is as important as building large theories: for example, we want to be able to build morphisms out of $\cn{CommutativeGroup}$ by combining compatible morphisms out of $\cn{Group}$ and $\cn{CommutativeMagma}$.

\paragraph{Flat Modules}
\mmt provides formal syntax for both structured modules and module expressions and defines their semantics via \textbf{flattening}, which defines for every module $M$ a flat module $\flt{M}$.

\textbf{Flat theories} are lists of declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The details can be found in \cite{rabe:howto:14}.

\textbf{Flat morphisms} from a theory $S$ to a theory $T$ are lists of assignments $c:=e$ where $c\in\dom{S}$ and $e\in\Exp{T}$.
Every morphism $M$ induces a \textbf{homomorphic extension} $M(-):\Exp{S}\to\Exp{T}$, which replaces every $c\in\dom{S}$ in an $S$-expression with the $T$-expression $e$ such that $c:=e$ in $M$.

Both theories and morphisms are subject to typing conditions that are not essential for our purposes here.
We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
This includes the preservation of truth via the propositions-as-types principle where $E$ is a proposition and $e$ its proof.

An \mmt \textbf{diagram} consists of a set of named structured module declarations and induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
Then flattening assigns
\begin{compactitem}
 \item to each $T\in\Thy$ the flat theory $\flt{T}$,
 \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
\end{compactitem}

@@

We start with the syntax for theories (which arises as a special case of the one given in \cite{RK:mmt:10}):

\begin{definition}[Theory]\label{def:theory}
A \textbf{theory} $T$ is of the form
\begin{grammar}
TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n\opt{:o}\opt{=o}     & \text{constant declaration}\\
c        & t?n                   & \text{qualified constant identifiers} \\
o        & c \alt \ldots         & \text{expressions built from constants} \\
T        & t                     & \text{theory expressions}
\end{grammar}

In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions only previously introduced constants.
We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
\end{definition}

At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding production for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
This makes it easy to incrementally add new declaration kinds and operators to the language.

%These theories are similar to OpenMath content dictionaries \cite{openmath}.
%In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
%\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
%But the interface always knows the qualified identifier and shows it, e.g., on hover.}
%Thus, symbols of the same name declared in different atomic theories are always distinguished.

We proceed accordingly for flat morphisms:

\begin{definition}[Morphism]\label{def:morphism}
A \textbf{morphism} is of the form
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
Ass    & c:=o                            & \text{assignment to symbol}\\
M      & m                               & \text{morphism expressions}
\end{grammar}
such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
\end{definition}

%Crucially, morphisms \textbf{preserve judgments}, in particular they map theorems to theorems.
%Formally, an atomic morphism $m:S\to T$ is \emph{well-formed} if
%\begin{compactitem}
%\item whenever $c$ has type $t$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o:m(t)$,
%\item whenever $c$ has definiens $d$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o\equiv m(d)$.
%\end{compactitem}
%Then we can show that for every $M\in\Mo{S}{T}$ if $\der_S o:o'$ (or $\der_S o\equiv o'$), then $\der_T M(o):M(o')$ (or $\der_T M(o)\equiv M(o')$).
%The details can be found in \cite{rabe:howto:14}.
%%In particular, this means that $M$ maps $S$-theorems to $T$-theorems.

Finally, we define diagrams as collections of modules:

\begin{definition}[Diagram]
A \textbf{diagram} is a list of theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
A diagram is well-formed if each atomic theory/morphism declaration has a unique name is well-formed relative to the diagram preceding it.
\end{definition}

We give some examples how to add modularity principles to the grammar:

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
\end{example}

\begin{example}[Union]\label{syn:union}
We extend the grammar with
\begin{grammar}
  T   & T\cup T & \text{union of theories} \\
  M   & M\cup M & \text{union of morphisms}
\end{grammar}
\end{example}

\begin{example}[Category of Theories]\label{syn:cat}
To obtain the category structure of theories and morphisms, we need identity and composition.
We add them to the grammar as follows:
\begin{grammar}
  M   & \id{T} & \text{identity morphism} \\
  M   & M;M    & \text{morphism composition}
\end{grammar}
\end{example}

For each example, we will define the semantics in the next section.

@@

\paragraph{Flattening}
Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $flt{Y}$.

For theories $T$, at this point, this induction only has the base case of a reference to a declared theory $t$.
Further cases arise when we add structuring declarations.

For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
\[\flt{t}=\flt{\Sigma}\]
where
 \[\flt{\cdot}=\es\]
 \[\flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
Here $\fltd{T}{D}$ is the flattening of declaration $d$ in the context of theory $T$.

At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{n[:E][=e]}=\{t?n[:E][=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{c:=e}=\{c:=o\}\]
\bigskip

The semantics serves two purposes:
firstly, it flattens every well-formed module into a flat one;
secondly, it determines which modules are well-formed at all.
It is desirable to define well-formedness independently of flattening.
However, in practice, it shows that implementations typically have to do almost the entire flattening to check well-formedness anyway.

Therefore, it is convenient to merge the two into one operation by simply making flattening a partial function: a module $X$ is \textbf{well-formed} iff $\flt{X}$ is defined.
\bigskip

\begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
Includes add new declarations $D$, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
We do this as follows
  \[\fltd{\Sigma}{\icl{S}} = \flt{S}\]
This has the effect of copying over all declarations from the included into the including theory.

Note that the definition of $\fltd{\Sigma}{n:E=e}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
Therefore, includes can never lead to name clashes.

Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening and well-formedness must be distinguished:
  \[\fltd{\sigma}{\icl{M}} = \flt{M}\]
under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
\end{example}

\begin{example}[Union (continued from Ex.~\ref{syn:union})]\label{sem:union}
We define the semantics of unions of theories as
\[\flt{(S\cup T)}=\flt{S}\cup\flt{T}\]
The handling of name clashes is the same as for includes:
Because the theories declared in a diagram have unique names and declarations in the flattening are qualified by theory names, union theories cannot have name clashes.
But we have to watch out when taking unions of morphisms:
We define $M_1\cup M_2\in \Mo{S_1\cup S_2}{T}$ if $M_i\in\Mo{S_i}{T}$ and $M_1$ and $M_2$ agree for every constant in $\dom{\flt{S_1}}\cap\dom{\flt{S_2}}$.
In that case, we put
\[\flt{(M\cup N)}=\flt{M}\cup\flt{N}\]
\end{example}

Finally, we define the semantics of the category structure:
\begin{example}[Category of Theories (continued from Ex.~\ref{syn:cat})]\label{sem:cat}
For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
\[\flt{\id{T}}=\{c:=c \;|\; c\in\dom{\flt{T}}\}\]

For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c\in\dom{\flt{R}}\}\]
\end{example}

%Includes between theories are always well-formed (assuming the included theory is).
%But includes between morphisms are more difficult -- we have to impose two restrictions.
%Firstly, an assignment $\icl{M}$ is well-formed in an atomic morphism $m:S\to T$ if $R\harr S$ and $M\in\Mo{R}{T}$.
%Its intended effect is that $m$ maps all $c\in\dom{R}$ to $M(c)$.
%That yields the second restriction: For any two morphisms $M_i\in\Mo{R_i}{T}$ included into $m$ and any theory $C$ satisfying $C\harr R_1$ and $C\harr R_2$, we require that $M_1|_C\equiv M_2|_C$.
%Thus, any two included morphisms must agree on their common domain $C$; that is necessary so that $\flt{m}$ does not contain multiple assignments for the same identifier.

\subsection{Semantic Relations}

For two theories $S,T\in\Thy$, we say that $S$ is \textbf{included} into $T$, written $S\harr T$, if $\flt{S}\sq\flt{T}$.
%Clearly, $m(o)=o$ for all $S$-objects $o$.
Thus, every declaration of $S$ is also available in $T$ and $\Exp{S}\sq\Exp{T}$.
This induces an inclusion morphism, which maps every $c\in\dom{S}$ to itself.
We do not introduce a name for that morphism and simply define $\id{S}\in\Mo{S}{T}$ whenever $S\harr T$ (which includes the special case $S=T$).
Moreover, if $S\harr T$ and $M\in\Mo{T}{U}$, we define the restriction $M|_S\in\Mo{S}{U}$ of $M$ to $S$ by $\id{S};M$.

@@

Correspondingly, we say that two morphisms $M,N\in\Mo{S}{T}$ are \textbf{equal}, written $M\equiv N$, if $\flt{M}=\flt{N}$.
In that case, $M(e)=N(e)$ for all expressions $e\in\Exp{S}$.

Clearly, $\equiv$ is an equivalence relation and $\harr$ is an order relation (if anti-symmetry is taken with respect to $\equiv$).

\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{rel:incl}
We have the following soundness theorem for includes: if a theory is declared as $t=\{\Sigma\}$ and $\Sigma$ contains $\icl{S}$, then $S\harr t$.

Note that in the absence of include declarations or unions, $s\harr t$ never holds because all $c\in\dom{s}$ are of the form $s?n$ and all $c\in\dom{t}$ are of the form $t?n$.
In fact, $\flt{s}$ and $\flt{t}$ unless $s=t$.
\end{example}

\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ is the universal morphism out of it.
\end{example}

\begin{example}[Category of Theories (continued from Ex.~\ref{sem:cat})]\label{rel:cat}
Identity and composition are congruent with respect to $\equiv$.
Moreover, composition is associative and identity neutral with respect to $\equiv$.
Thus, each diagram indeed yields the structure of a category with objects $\Thy$ and arrows $\Mo{S}{T}$.
\end{example}

### 2018-09-26 wadtabstract.tex
rewritten Sect. 1+2

@@

>>>>

\documentclass[orivec]{llncs}

\usepackage{graphicx,url,wrapfig,amsmath,amssymb}

\usepackage[show]{ed} %for ednotes

\usepackage{sembracket}
\usepackage{mytikz}
\usepackage{basics}
%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
\usepackage{local}

\setcounter{tocdepth}{3}
\usepackage[bookmarks,linkcolor=red,citecolor=blue,urlcolor=gray,colorlinks,breaklinks,bookmarksopen,bookmarksnumbered]{hyperref}

\setlength{\hfuzz}{3pt}
\hbadness=10001 %make box warning less strict

\begin{document}
\title{Structuring Theories with Implicit Morphisms}
\author{Florian Rabe$^{1,2}$ \and Dennis M\"uller$^2$}
\institute{LRI Paris \and FAU Erlangen-Nuremberg}
\maketitle

\begin{abstract}
We introduce \emph{implicit} morphisms as a concept in formal systems based on theories and theory morphisms.
The idea is that there may be at most one implicit morphism from a theory $S$ to a theory $T$, and if $S$-expressions are used in $T$ their semantics is obtained by automatically inserting the implicit morphism.
The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
\end{abstract}

@@

\bibliographystyle{alpha}
\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}
\end{document}
